---
output: 
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
geometry: margin=2.54cm
title: "Comparison of Effects of COVID-19 on Counties With and Without Coal Mines"
subtitle: "https://github.com/tnh2/ENV872_Project"
author: "Thomas Hancock"
fontsize: 12pt
mainfont: Times New Roman

---

\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	echo = FALSE,
	include = TRUE
)
# Set your working directory
getwd()

# Load your packages
library(tidyverse)
library(lubridate)
library(knitr)
library(dplyr)
library(viridis)
library(RColorBrewer)
library(here)
library(sf)
library(ggmap)
library(cowplot)


# Set your ggplot theme
mytheme <- theme_classic() +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "bottom")
theme_set(mytheme)

# Load your datasets

```


# Rationale and Research Questions

The purpose of this analysis is to explore whether counties with coal mines are impacted differently by COVID-19. With the novel coronavirus rapidly spreading through the United States, different locations are likely to be affected in different ways. In order to perform the analysis, I combined datasets containing information about coal mine locations, populations, and COVID-19 cases and deaths at the county level. Because other factors such as socioeconomic levels, population densities, and even government responses to the COVID-19 crisis vary widely across states, I decided that it was best to compare counties within states rather than performing analysis on the entire country. In particular, I focused on Pennsylvania and West Virginia. Pennsylvania has had a relatively large number of confirmed COVID-19 cases, meaning there is a reasonable amount of data to analyze. West Virginia has had relatively few confirmed cases, but I included it as a comparison to Pennsylvania since West Virginia coal mines have continued operation whereas Pennsylvania mines were closed in March^1^.

I am particularly interested in whether or not there will be a difference in COVID-19 fatality (the percentage of people who have the disease that die from it) between the counties with coal mines and those without. The rational behind this exploration is that COVID-19 seems to be particularly lethal among people with existing conditions, especially respiratory issues. In turn, coal miners frequently suffer respiratory issues from the large amonts of dust that they encounter in their jobs. As such, I wondered if there might be a higher fatality rate in coal mining counties than non-coal mining counties (as a proxy for fatality amongst coal miners versus others). Likewise, coal mining involves sustained close-quarters operation and minimal opportunity for sanitization. These factors make coal mines a potential hotbed for COVID-19 spread, leading me to compare counties in Pennsylvania where the mines shut down in response to the outbreak against counties in West Virginia where mines continued operation.

My research questions are:

1. Do counties with coal mines have different fatality rates than similar counties without coal mines in the same state?

2. Do counties with coal mines in Pennsylvania have different fatality rates than West Virgnia counties with coal mines?


\newpage

# Dataset Information

To perform my analysis, I gathered together five datasets: US Census state and county boundaries (as shapefiles), US Census population data for US counties, coal mine locations from the US Department of Energy (as a shapefile), and county-level COVID-19 case and death counts from the Johns Hopkins University Center for Systems Science and Engineering (CSSE). The population data from the US Census Bureau are from 2019 estimates. The coal mine information inlcudes all operating surface and underground mines in the United States. While the dataset includes information on the mine type and output, I only used the location data for this analysis. The COVID-19 data from Johns Hopkins includes confirmed COVID-19 cases and reported deaths by county as of April 23rd, 2020.

I started wrangling my data by finding which counties had coal mines in them using the geospatial analysis capabilities of the sf package in R. I then joined the COVID-19 data with the census population data using the county FIPS codes and calculated metrics for COVID-19 case rates (per 100,000 people), fatality rate (deaths per confirmed COVID-19 case), and death rates per 100,000 people in the county. This data was then combined with the data above identifying whether a county had coal mines or not to complete the data joining. From this country-wide dataset, I filtered out the data specific to the two states I focused on for this study (Pennsylvania and West Virginia). The important data included in these data sets are outlined in Table 1 below.


Table: Important Data Fields

Field Name      | Description                  | Units 
----------------|------------------------------|--------
STNAME          | State Name                   | 
CTYNAME         | County Name                  |
POPESTIMATE2019 | 2019 Estimated Population    | People 
FIPS            | 5-digit county identifier    | 
Confirmed       | Confirmed COVID-19 cases     | Total Cases
Deaths          | Reported COVID-19 deaths     | Total Deaths
case_rate       | Rate of COVID-19 cases       | Cases per 100,000 people
death_rate_conf | Fatality                    | Deaths per confirmed case
death_rate_pop  | Deaths rate in population    | Deaths per 100,000 people
Mines.Label     | Whether the County has mines | Mines or No Mines

\newpage

# Exploratory Analysis 

## Pennsylvania Data
Turning my attention to Pennsylvania, I started by creating a histogram of the fatality rates of the coal counties and the non-coal counties. The first histogram I created is shown in Figure 1 below. 

```{r PA Histogram1, echo=FALSE, fig.cap="Histogram of Fatality in Pennsylvania counties", message=FALSE, warning=FALSE}
# Load PA data
PA.pop <- read.csv(here("Data", "Processed", "PA.Pop_processed.csv"))

ggplot(PA.pop) +
  geom_histogram(aes(x = death_rate_conf, fill = Mines.Label), bins = 18) +
  labs(x = "Fatality", y = "Count", fill = "") +
  scale_fill_viridis_d(option = "inferno", end = 0.75)

```

Two things are immediately evident: the data are not normally distributed, and there are some strange cases where the fatality is 1 (i.e., every confirmed case of COVID-19 resulted in a death). Based on general knowledge that COVID-19 is not likely to be lethal enough to kill even a majority of patients, we can assume that these counties had such limited testing that only the mortalities were tested. For the sake of this analysis, we will discard those counties and any others that have a fatality rate greater than 50%. We will assume that the remaining counties had testing that was sufficiently widespread to provide a reasonable approximation of the fatality. This new dataset yields the histogram in Figure 2. The distribution is still not normal, but it has a smoother shape for both the counties with mines and those without. There are still a few counties with suspiciously higher fatality rates, but without further data, there is not enough reason to remove them.

```{r PA Histogram2, echo=FALSE, fig.cap="Histogram of Fatality in Pennsylvania counties (outliers removed)", message=FALSE, warning=FALSE}
# Filter for counties with death rate > 0.5
PA.pop_testing <- filter(PA.pop, death_rate_conf < 0.5)

ggplot(PA.pop_testing) +
  geom_histogram(aes(x = death_rate_conf, fill = Mines.Label), bins = 18) +
  labs(x = "Fatality", y = "Count", fill = "") +
  scale_fill_viridis_d(option = "inferno", end = 0.75)
```

I decided to explore the potential relationship between fatality and the amount of testing by plotting fatality against the number of confirmed cases per 100,000 people in each county. The results are presented in Figure 3. Even without the aforementioned counties that only tested fatal cases, there appears to be a slight decreasing trend in fatality as the confirmed case rate increases. However, this trend is not statistically significant (linear regression; p-value = 0.395; F-statistic = 0.7344; degrees of freedom: 1 and 64).

```{r PA Fatality vs Case Rate, echo=FALSE, fig.cap=" Fatality vs confirmed COVID-19 case rate for Pennsylvanian counties with and without coal mines.", message=FALSE, warning=FALSE}

ggplot(PA.pop_testing, aes(x = case_rate, y = death_rate_conf, color = Mines.Label)) +
  geom_point(size  = 2) +
  labs(x = "Confirmed Cases per 100,000 People", y = "Fatality", 
       color = "") +
  scale_color_viridis_d(option = "inferno", end = 0.75)

fatality.case_rate <- lm(data = PA.pop_testing, death_rate_conf ~ case_rate)
```

## West Virginia Data
As mentioned earlier, I am also interested in seeing if there is a difference between West Virginia, where coal mines stayed open despite the pandemic, and Pennsylvania, where mines were closed. In Figure 4, I plotted histograms for the West Virginia data, similar to what was done for Pennsylvania. From these histograms, we can see that there were a few West Virignia counties that tested mostly fatal cases of COVID-19, so it is reasonable to perform the same removal of ouliers with fatality rates above 50%. Likewise, the West Virginia data has a very non-normal distribution.

```{r WV Histograms, echo=FALSE, fig.cap="Histogram of Fatality in West Virginia Counties (a) with outliers and (b) without outliers.", message=FALSE, warning=FALSE}

# Load West Virginia data
WV.pop <- read.csv(here("Data", "Processed", "WV.Pop_processed.csv"))

# Filter for counties with death rate > 0.5
WV.pop_testing <- filter(WV.pop, death_rate_conf < 0.5)

WV.hist1 <- ggplot(WV.pop) +
  geom_histogram(aes(x = death_rate_conf, fill = Mines.Label), bins = 20) +
  labs(x = "Fatality", y = "Count", fill = "NULL") +
  theme(legend.position = "none") +
  scale_fill_viridis_d(option = "inferno", end = 0.75)

WV.hist2 <- ggplot(WV.pop_testing) +
  geom_histogram(aes(x = death_rate_conf, fill = Mines.Label), bins = 15) +
  labs(x = "Fatality", y = "Count", fill = "") +
  scale_fill_viridis_d(option = "inferno", end = 0.75)

plot_grid(WV.hist1, WV.hist2, nrow = 2, align = 'h', labels = c("(a)", "(b)"), label_size = 12)

```


Plotting the fatality against the confirmed case rate for West Virginia counties, as shown in Figure 5, reveals little structure to the data. Perhaps the most useful conclusion that can come from this exploration of the data is that West Virginia has a rather low reported case rate.

```{r WV Fatality vs Case Rate, echo=FALSE, fig.cap=" Fatality vs confirmed COVID-19 case rate for West Virginian counties with and without coal mines.", message=FALSE, warning=FALSE}

ggplot(WV.pop_testing, aes(x = case_rate, y = death_rate_conf, color = Mines.Label)) +
  geom_point(size  = 2) +
  labs(x = "Confirmed Cases per 100,000 People", y = "Fatality", 
       color = "") +
  scale_color_viridis_d(option = "inferno", end = 0.75)

fatality.case_rate <- lm(data = WV.pop_testing, death_rate_conf ~ case_rate)
```


\newpage

# Analysis

## Question 1: Do counties with coal mines have different fatality rates than similar counties without coal mines in the same state?

To answer the first question comparing coal mining counties in Pennsylvania to non-mining counties, I performed a non-parametric Wilcoxon rank sum test. I chose to use a non-parametric test because the data are not distributed normally (as seen in the Exploration section). The Wilcoxon test uses a ranking system to see if two data sets come from the same distribution. According to this test, there is no significant difference between the fatality rates in coal mining counties versus non-mine counties (Wilcoxon test; p = 0.7516; W = 536.5). To visualize this result, I made a boxplot for both sets of data (fatality in each type of county). This boxplot is shown in Figure 6. The similarity between the two boxplots reinforces the fact that the two sets of data are similar.

```{r Wilcoxon_PA, message=FALSE, warning=FALSE, include=FALSE}
# Run Kruskal-Wallis test
Wilcoxon.PA <- wilcox.test(filter(PA.pop_testing, Mines.Label == "Mines")$death_rate_conf,
                      filter(PA.pop_testing, Mines.Label == "No Mines")$death_rate_conf)
```

```{r PA_Boxplots, fig.cap="Boxplots of fatality in Pennsylvania counties.", message=FALSE, warning=FALSE}
ggplot(PA.pop_testing) +
  geom_boxplot(aes(x = Mines.Label, y = death_rate_conf)) +
  labs(x = "", y = "Fatality")
```

## Question 2: Do counties with coal mines in Pennsylvania have different fatality rates than West Virgnia counties with coal mines?

I also used a Wilcoxon test for the second question. As with the first question, the West Virginia and Pennsylvania mine counties have non-normal distributions. Once again, there is no significant difference between the fatality cases in the mining counties of the two states (Wilcoxon Test; p = 0.0905; W = 306.5). Boxplots of the fatality rates in the two states also show that they are not significantly different (Figure 7).

```{r Wilcoxon_PA-WV, message=FALSE, warning=FALSE, include=FALSE}
# Run Kruskal-Wallis test
Wilcoxon.PA.WV <- wilcox.test(filter(PA.pop_testing, Mines.Label == "Mines")$death_rate_conf,
                      filter(WV.pop_testing, Mines.Label == "Mines")$death_rate_conf)
```

```{r PA-WV_Boxplots, fig.cap="Boxplots of fatality in mining counties in Pennsylvania and West Virginia.", message=FALSE, warning=FALSE}
PA.WV.mines <- rbind(PA.pop_testing, WV.pop_testing)
ggplot() +
  geom_boxplot(data = filter(PA.WV.mines, 
                             Mines.Label == "Mines"), aes (x = STNAME, y = death_rate_conf)) +
  labs(x = "", y = "Fatality")

```

\newpage

# Summary and Conclusions

Based on the analyses explained earlier, there does not appear to be a statistically significant difference in COVID-19 effects on counties with and without coal mines (in Pennsylvania and West Virginia, at least). Similarly, the fact that Pennsylvania closed their coal mines due to the pandemic and West Virginia did not does not seem to have led to a higher fatality rate in West Virginia. My initial thought was that counties with coal mines might have higher fatality rates because of the more severe outcomes associated with COVID-19 in patients with pre-existing conditions and the prevalence of pulmonary disease among coal miners. However, there are a number of factors that could account for this data not yielding these results.

One possibility is that it is too early in the US pandemic to see enough cases to accurately assess mortality. Especially in West Virginia and more rural parts of Pennsylvania, there has not been thorough testing, so the data might not have enough statistical power. Moreover, the virus might not have infected enough people in these regions yet to be able to compare the outcomes. It is possible that, as the virus spreads, more cases (and, subsequently, more fatality data) will become available. Perhaps by the end of this summer, a similar analysis might have a different outcome. Conversely, it is possible that the virus will not impact these regions as much as expected. A number of theories have been proposed for why this low incidence might be observed, including the fact that these smaller communities tend to be more spread out (so the virus spreads slower) or that the virus has actually been in West Virginia longer than is commonly thought (potentially since December). Even if there is truth to these ideas, it seems very likely that a major limitation of this study is limited testing, which results in a lack of data.

Another factor that might have led to these negative results is that the number of mining community members who are actually miners is often quite low (especially at the county level). As such, the number of people in the county who are actually compromised because of the presence of the mine is relatively low. Socioeconomic factors, or even potential vulnerabilities due to pollution from the mine, can cross county lines to other non-mining counties, which would lessen the difference seen when comparing the two.

Nonetheless, it is interesting to see that there is no appreciable difference bewteen mining and non-mining counties when it comes to COVID-19 fatalities. Further work could be done to identify factors that can contribute to differences in fatality (including prevalence of testing, air quality issues, and socioeconomic factors, among others). It will also be interesting to see how the COVID-19 situation evolves, and similar analyses may be warranted in a number of months when more data is available.

\newpage

# References
1. Smith, J. (2020, April 5). In West Virginia, experts say coal mines could be massive spreading ground for COVID-19. Mines across the state are still operating normally. *Times West Virginia*, Retrieved from www.timeswv.com
